---
title: "data_analysis_for_McManus"
author: "Siqi"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r }
# Load necessary libraries jiayi
set.seed(123)
library(ggplot2)
library(splines)
library(minpack.lm)
#install.packages("pracma") 
library(pracma)
#install.packages("nls2")
library(nls2)


```

## Including Plots

Data Cleaning and Plotting

```{r  }
# Load and process the  
library(readxl)
Data_McManus2_WT_P2L_AAVCremCherry <- read_excel("/Users/angelzhang/Downloads/CircadianProject/Data_McManus2_WT_P2L_AAVCremCherry.xlsx")


df_MC <- as.data.frame(Data_McManus2_WT_P2L_AAVCremCherry)

# Rename columns for clarity
colnames(df_MC) <- c("Time", "X1", "X2", "X3", "X4")

# Remove outliers where Time is between 100 and 100.1
df_MC <- df_MC[!(df_MC$Time >= 100 & df_MC$Time <= 100.1), ]

# Plot each species
ggplot(df_MC, aes(x = Time)) +
  geom_line(aes(y = X1, color = "X1")) +
  geom_line(aes(y = X2, color = "X2")) +
  geom_line(aes(y = X3, color = "X3")) +
  geom_line(aes(y = X4, color = "X4")) +
  labs(title = "Data without Outliers", x = "Time", y = "X(t_i)") +
  scale_color_manual(values = c("X1" = "blue", "X2" = "green", "X3" = "red", "X4" = "purple")) +
  theme_minimal()

```

```{r}

spline_X1 <- predict(smooth.spline(df_MC$Time, df_MC$X1),    spar =)
spline_X2 <- predict(smooth.spline(df_MC$Time, df_MC$X2))
spline_X3 <- predict(smooth.spline(df_MC$Time, df_MC$X3))
spline_X4 <- predict(smooth.spline(df_MC$Time,df_MC$X4))

# Plot the spline-smoothed data
plot(df_MC$Time, spline_X1$y, type = "l", col = "red", lwd = 2, 
     xlab = "Time", ylab = "Values", main = " Splines", 
     ylim = range(c(spline_X1$y, spline_X2$y, spline_X3$y, spline_X4)))
lines(df_MC$Time, spline_X2$y, col = "blue", lwd = 2)
lines(df_MC$Time, spline_X3$y, col = "green", lwd = 2)
lines(df_MC$Time, spline_X4$y, col = "purple", lwd = 2)

legend("topright", legend = c("spline_X1", "spline_X2", "spline_X3", "spline_X4"), 
       col = c("red", "blue", "green", "purple"), lwd = 2)


# Plot the spline-smoothed data
spline_X1 <- predict(smooth.spline(df_MC$Time, df_MC$X1, spar=1)) #black line in 2nd #plot（average for X1）

plot(df_MC$Time, df_MC$X1, type = "l", col = "yellow", lwd = 2, 
     xlab = "Time", ylab = "Values", main = " Splines", 
     ylim = range(c(spline_X1$y, spline_X2$y, spline_X3$y, spline_X4)))
lines(df_MC$Time, spline_X1$y)

plot(df_MC$Time, df_MC$X1-spline_X1$y, type = "l", col = "yellow", lwd = 2, 
     xlab = "Time", ylab = "Values", main = " Detrended Splines")



```
```{r}
x <- 1:length(df_MC$Time)  
y <- df_MC$X1-spline_X1$y    

maxima <- findpeaks(df_MC$X1-spline_X1$y, minpeakheight=0)  
minima <- findpeaks(-df_MC$X1-spline_X1$y, minpeakheight=0) 

x_max <- maxima[,2]  
y_max <- maxima[,1]  


fit <- nls(y_max ~ a * exp(-b * x_max) + c, start = list(a = max(y_max), b = 0.01, c = min(y_max)))

summary(fit)

a <- coef(fit)["a"]
b <- coef(fit)["b"]
c <- coef(fit)["c"]

plot(x, y, type="l", col="yellow", main="Detrended Splines with Exponential Fit")
lines(x_max, predict(fit, list(x_max = x_max)), col="red", lwd=2)  #

```



Simulate Data Based on the Fitted Model

```{r}
```


```{r}
t <- seq(0, 100, by = 0.1)

model_function <- function(t, A, gamma, omega, phi, y_shift) {
  A * exp(-gamma * t ) * cos(omega * t + phi) + y_shift
}

# using estimated parameters to def simulating function
simulate_data <- function(t, params, sigma) {
  predicted <- model_function(t, params[1], params[2], params[3], params[4], params[5])
  noise <- rnorm(length(t), mean = 0, sd = sigma)  #using sigma0 to produce noise 
  simulated_data <- predicted + noise
  return(simulated_data)
}
sigma0 = 10
param0=c(1000, 0.025, 0.9, 0,0)
simulated_data <- simulate_data(t, param0, sigma0 )
```


```{r}
plot(t, simulated_data, type = "l", col = "blue", main = "Simulated Data", xlab = "Time", ylab = "Simulated Values")
```





##Task1
##generate simulating data 1 time

```{r}


true_params <- c(A = 1000, gamma = 0.025, omega = 0.9, phi = 0, y_shift = 0)

sigma0 <- 10
t <- seq(0, 100, by = 0.1)
set.seed(123)
y_sim <- model_function(t, true_params[1], true_params[2], true_params[3], true_params[4], true_params[5]) + 
         rnorm(length(t), mean = 0, sd = sigma0)

# default algorithm in nls: Gauss-Newton
fit_nls <- nls(
  y_sim ~ model_function(t, A, gamma, omega, phi, y_shift),
  start = list(A = 950, gamma = 0.03, omega = 0.85, phi = 0.1, y_shift = 0), #initial value close to true ones
  data = data.frame(t = t, y_sim = y_sim)
)
# 
summary(fit_nls)
coef(fit_nls)
```

##Task 2 find peaks
```{r}
t <- seq(0, 100, by = 0.1)
# find peaks
peaks <- findpeaks(y_sim, minpeakheight = 100, minpeakdistance = 15) 

# peaks[, 2] :index of peaks in (1~1000)
#t[ peaks[, 2]]: index of peaks in(1~100)
t_peaks <- t[peaks[, 2]] 

y_peaks <- peaks[, 1]

#
plot(t, y_sim, type = "l", col = "blue", main = "Simulated Data with Peaks", xlab = "Time", ylab = "Simulated Values")
points(t_peaks, y_peaks, col = "red", pch = 19)
```
##Finding initial parameters:A and gamma
```{r}
# transform into log
log_y_peaks <- log(y_peaks)

# logy = -gamma*t + logA
fit_linear <- lm(log_y_peaks ~ t_peaks) 

summary(fit_linear)
print("coef:")
coef(fit_linear)
coef(fit_linear)[1]
print("coef:")

# get log(A)  and  gamma
log_A <- coef(fit_linear)[1]

init_gamma <- -coef(fit_linear)[2] #Extract gamma from the linear fit

cat("log(A) =", log_A, "\n")
cat("gamma =", init_gamma, "\n")
init_A=exp(log_A)
print(init_A)
```

## to see whether the initial params A and gamma is fitted or not
```{r}
#Plot the simulated data
plot(t, y_sim, type = "l", col = "blue", main = "Simulated Data with Fitted Exponential", xlab = "Time", ylab = "Simulated Values")

# Calculate the fitted exponential curve using estimated A and gamma from the linear model
init_A <- exp(log_A)  # Extract log_A from the linear model

# Add the fitted exponential curve
curve(init_A * exp(-init_gamma * x), from = 0, to = max(t), col = "green", add = TRUE, lwd = 2)
#  Add a legend 
legend("topright", legend = c("Simulated Data", "Fitted Exponential"), col = c("blue", "green"), lty = 1)

```

##find initial value：omega
```{r}
# Extract the time points of the first two consecutive maxima from t_peaks
t_max1 <- t_peaks[1]  # First maximum
t_max2 <- t_peaks[2]  # Second maximum
t_max3 <- t_peaks[3] 
t_max4 <- t_peaks[4] 
# Calculate the time difference between two consecutive maxima
Delta <- t_max2 - t_max1
Delta 
print(t_max3-t_max2)
print(t_max4 - t_max3)
# Estimate omega (tau) using the formula 2*pi/Delta
init_omega <- 2 * pi / Delta

# Output the result
cat("initial omega (tau) =", init_omega, "\n")
cat("Time difference between consecutive maxima (Delta) =", Delta, "\n")

```
we assume period delta keeps same. now our initial parameters are[A = 968.7701,gamma = 0.02295209 ,omega = 0.9377889 , phi = 0, y_shift = 0]

```{r}
library(nlme)
  
fit_nls <- nls(
  y_sim ~ model_function(t, A, gamma, omega, phi, y_shift),
  start = list(A = init_A, gamma = init_gamma, omega = init_omega, phi = 0.1, y_shift = 0),
  data = data.frame(t = t, y_sim = y_sim)
)
fit_nls

```


###Tasks 3
```{r}

#library(stats)   

simulate_and_estimate <- function(true_params, sigma0 = 10) {
  
  #simulate data
  t <- seq(0, 100, by = 0.1)
  
  
  model_function <- function(t, A, gamma, omega, phi, y_shift) {
    A * exp(-gamma * t) * cos(omega * t + phi) + y_shift
  }
  
  
  y_sim <- model_function(t, true_params[1], true_params[2], true_params[3], true_params[4], true_params[5]) + 
    rnorm(length(t), mean = 0, sd = sigma0)
  
  # find initial value
  peaks <- findpeaks(y_sim, minpeakheight = 100, minpeakdistance = 15)
  t_peaks <- t[peaks[, 2]]
  y_peaks <- peaks[, 1]
  
  log_y_peaks <- log(y_peaks)
  fit_linear <- lm(log_y_peaks ~ t_peaks)
  log_A <- coef(fit_linear)[1]
  init_gamma <- -coef(fit_linear)[2]
  init_A <- exp(log_A)
  
  
  Delta <- t_peaks[2] - t_peaks[1]
  init_omega <- 2 * pi / Delta
  
    # find estimated parameters
  fit_nls <- try(nls(
    y_sim ~ model_function(t, A, gamma, omega, phi, y_shift),
    start = list(A = init_A, gamma = init_gamma, omega = init_omega, phi = 0.1, y_shift = 0),
    data = data.frame(t = t, y_sim = y_sim)
  ), silent = TRUE)
  
  # check if fitted successful
  if (class(fit_nls) == "try-error") {
    return(NULL)  
  }
  # find estmate parameters and standard errors
  param_estimates <- coef(fit_nls)
  cov_matrix <- vcov(fit_nls)
  
  standard_errors <- sqrt(diag(cov_matrix))
  z_value <- qnorm(0.995)
  conf_intervals <- cbind(
    param_estimates - z_value * standard_errors,
    param_estimates + z_value * standard_errors
  )
  return(list(estimates = param_estimates, conf_intervals = conf_intervals))
}
# test function
true_params <- c(A = 1000, gamma = 0.025, omega = 0.9, phi = 0, y_shift = 0)
result <- simulate_and_estimate(true_params)

result


```
##Task 3: repeat simulations
```{r}
# simulat
n_simulations <- 100

# 
results <- vector("list", n_simulations)
# 
for (i in 1:n_simulations) {
  results[[i]] <- simulate_and_estimate(true_params)
}
results <- Filter(Negate(is.null), results)
null_count <- sum(sapply(results, is.null))
cat("Number of unsuccessful fits (NULL results):", null_count, "\n")
# Initialize coverage counters, one counter for each parameter
coverage_counts <- rep(0, length(true_params))

# Iterate through each simulation result to check coverage.
for (i in 1:length(results)) {
  ci <- results[[i]]$conf_intervals
  for (j in 1:length(true_params)) {
    # Check if it falls within the confidence interval.
    if (true_params[j] >= ci[j, 1] && true_params[j] <= ci[j, 2]) {
      coverage_counts[j] <- coverage_counts[j] + 1
    }
  }
}
#calculate coverage
coverage_proportions <- coverage_counts / n_simulations
# print unsuccessful fits


cat("Coverage for A:", coverage_proportions[1], "\n")
cat("Coverage for gamma:", coverage_proportions[2], "\n")
cat("Coverage for omega:", coverage_proportions[3], "\n")
cat("Coverage for phi:", coverage_proportions[4], "\n")
cat("Coverage for y_shift:", coverage_proportions[5], "\n")

```
`


```{r}
library(tidyverse) 
library(dplyr)
library(tidyr)

# Extract valid results (remove NULL values);
valid_results <- Filter(Negate(is.null), results)

# Extract parameter estimates into a data frame
estimate_data <- do.call(rbind, lapply(valid_results, function(res) {
  as.data.frame(t(res$estimates))
}))


colnames(estimate_data) <- c("A", "gamma", "omega", "phi", "y_shift")

estimate_long <- estimate_data %>%
  pivot_longer(cols = everything(), names_to = "Parameter", values_to = "Estimate")

#install.packages("gridExtra")
library(gridExtra) # arange multiple plots

# Create separate boxplots for each parameter with suitable y-axis ranges;
plot_A <- ggplot(estimate_long %>% filter(Parameter == "A"), 
                 aes(x = Parameter, y = Estimate)) +
  geom_boxplot(fill = "skyblue", color = "darkblue") +
  labs(title = "A", y = "Estimate") +
  theme_minimal() +
  ylim(995, 1006)

plot_gamma <- ggplot(estimate_long %>% filter(Parameter == "gamma"), 
                     aes(x = Parameter, y = Estimate)) +
  geom_boxplot(fill = "lightgreen", color = "darkgreen") +
  labs(title = "gamma", y = "Estimate") +
  theme_minimal() +
  ylim(0.02475, 0.02525)

plot_omega <- ggplot(estimate_long %>% filter(Parameter == "omega"), 
                     aes(x = Parameter, y = Estimate)) +
  geom_boxplot(fill = "lightblue", color = "blue") +
  labs(title = "omega", y = "Estimate") +
  theme_minimal() +
  ylim(0.8995, 0.9005)

plot_phi <- ggplot(estimate_long %>% filter(Parameter == "phi"), 
                   aes(x = Parameter, y = Estimate)) +
  geom_boxplot(fill = "pink", color = "red") +
  labs(title = "phi", y = "Estimate") +
  theme_minimal() +
  ylim(min(estimate_data$phi) * 0.9, max(estimate_data$phi) * 1.1)

plot_y_shift <- ggplot(estimate_long %>% filter(Parameter == "y_shift"), 
                       aes(x = Parameter, y = Estimate)) +
  geom_boxplot(fill = "yellow", color = "orange") +
  labs(title = "y_shift", y = "Estimate") +
  theme_minimal() +
  ylim(min(estimate_data$y_shift) * 0.9, max(estimate_data$y_shift) * 1.1)

# Arrange all subplots into a single large plot.
grid.arrange(plot_A, plot_gamma, plot_omega, plot_phi, plot_y_shift, ncol = 3)
```
This plot indicates that most of the estimated parameters obtained from 100 simulations are close to the true parameters.

##Task 4 

finding initial parameters on asymmetric osillator model(on initial meeting file)
```{r}
time <- df_MC$Time
X1 <- df_MC$X1
# FInd X1 valleys and peaks
peaks <- findpeaks(X1, minpeakheight = 600, minpeakdistance = 20)
peaks<-peaks[-6, ]
t_peaks <- time[peaks[, 2]]
y_peaks<- peaks[, 1]
#delete outlier

valleys <- findpeaks(-X1, minpeakheight = -500, minpeakdistance = 50)

t_valleys <- time[valleys[, 2]]
y_valleys <- -valleys[, 1]  # restore 

# 
df_peaks <- data.frame(Time = t_peaks, Value = y_peaks)
df_valleys <- data.frame(Time = t_valleys, Value = y_valleys)


ggplot(df_MC, aes(x = Time, y = X1)) +
  geom_line(color = "blue") +
  geom_point(data = df_peaks, aes(x = Time, y = Value), color = "red", size = 2, shape = 17) +  
  geom_point(data = df_valleys, aes(x = Time, y = Value), color = "green", size = 2, shape = 18) + 
  labs(title = "X1 Data with Peaks and Valleys",
       x = "Time",
       y = "X1(t)") +
  theme_minimal()

```



```{r}
log_y_peaks <- log(y_peaks)
log_y_valleys <- log(y_valleys)

fit_peaks <- lm(log_y_peaks ~ t_peaks)
fit_valleys <- lm(log_y_valleys ~ t_valleys)

eta_max <- coef(fit_peaks)[2]  # η_max
eta_min <- coef(fit_valleys)[2]  # η_min

# estimate A_max and A_min
A_max <- exp(coef(fit_peaks)[1])
A_min <- exp(coef(fit_valleys)[1])


Delta <- t_peaks[2] - t_peaks[1]
tau <- 2 * pi / Delta

#assume phi = 0
phi <- 0

# print
cat("Initial A_max:", A_max, "\n")
cat("Initial A_min:", A_min, "\n")
cat("Initial eta_max:", eta_max, "\n")-
cat("Initial eta_min:", eta_min, "\n")
cat("Initial tau:", tau, "\n")
cat("Initial phi:", phi, "\n")

```
##plot fitted curve to compare
```{r}

max_exp_curve <- A_max * exp(eta_max * time)
min_exp_curve <- A_min * exp(eta_min * time)
# 
df_peaks <- data.frame(Time = t_peaks, Value = y_peaks)
df_valleys <- data.frame(Time = t_valleys, Value = y_valleys)
df_max_exp_curve <- data.frame(Time = time, ExponentialCurve = max_exp_curve)
df_min_exp_curve <- data.frame(Time = time, ExponentialCurve = min_exp_curve)
# 
ggplot(df_MC, aes(x = Time, y = X1)) +
  geom_line(color = "blue", size = 1) +  # 
  geom_line(data = df_max_exp_curve, aes(x = Time, y = ExponentialCurve), color = "purple", size = 1, linetype = "dashed") +  # expontential curve
  geom_line(data = df_min_exp_curve, aes(x = Time, y = ExponentialCurve), color = "yellow", size = 1, linetype = "dashed") +  # expontential curve
  geom_point(data = df_peaks, aes(x = Time, y = Value), color = "red", size = 2, shape = 17) +  #
  geom_point(data = df_valleys, aes(x = Time, y = Value), color = "green", size = 2, shape = 18) +
  labs(title = "X1 Data with Exponential Curve Fitting",
       x = "Time",
       y = "X1(t)") +
  theme_minimal() +
  scale_y_continuous(limits = c(0, max(X1) * 1.1))  #


```
using lm(),I cannot get a good fit, 
so I try to use nls to find initial A and gamma：
```{r}


nls_fit <- nls(y_peaks ~ A_max * exp(eta_max * t_peaks), 
               start = list(A_max = max(y_peaks), eta_max = -0.01), 
               control = nls.control(maxiter = 100))


A_max <- coef(nls_fit)["A_max"]
eta_max <- coef(nls_fit)["eta_max"]

# generate curve
exp_curve <- A_max * exp(eta_max * time)


df_exp_curve <- data.frame(Time = time, ExponentialCurve = exp_curve)
ggplot(df_MC, aes(x = Time, y = X1)) +
  geom_line(color = "blue", size = 1) +
  geom_line(data = df_exp_curve, aes(x = Time, y = ExponentialCurve), color = "purple", size = 1, linetype = "dashed") +
  geom_point(data = df_peaks, aes(x = Time, y = Value), color = "red", size = 2, shape = 17) +
  labs(title = "X1 Data with Improved Exponential Curve Fitting",
       x = "Time",
       y = "X1(t)") +
  theme_minimal() +
  scale_y_continuous(limits = c(0, max(X1) * 1.1))

```
still not good fit，

try to fit in 2 segments time before 100 and time after 100
```{r}
#  split data for early(t<100) and late segments(t>=100)
time_early <- time[time < 100]
time_late <- time[time >= 100]

y_peaks_early <- y_peaks[t_peaks < 100]
t_peaks_early <- t_peaks[t_peaks < 100]
y_valleys_early <- y_valleys[t_valleys < 100]
t_valleys_early <- t_valleys[t_valleys < 100]

y_peaks_late <- y_peaks[t_peaks >= 100]
t_peaks_late <- t_peaks[t_peaks >= 100]
y_valleys_late <- y_valleys[t_valleys >= 100]
t_valleys_late <- t_valleys[t_valleys >= 100]

# exponential fit when t <100
log_y_peaks_early <- log(y_peaks_early)
log_y_valleys_early <- log(y_valleys_early)

fit_peaks_early <- lm(log_y_peaks_early ~ t_peaks_early)
fit_valleys_early <- lm(log_y_valleys_early ~ t_valleys_early)

A_max_early <- exp(coef(fit_peaks_early)[1])
eta_max_early <- -coef(fit_peaks_early)[2]
A_min_early <- exp(coef(fit_valleys_early)[1])
eta_min_early <- -coef(fit_valleys_early)[2]

# exponential fit when t > 100
log_y_peaks_late <- log(y_peaks_late)
log_y_valleys_late <- log(y_valleys_late)

fit_peaks_late <- lm(log_y_peaks_late ~ t_peaks_late)
fit_valleys_late <- lm(log_y_valleys_late ~ t_valleys_late)

A_max_late <- exp(coef(fit_peaks_late)[1])
eta_max_late <- -coef(fit_peaks_late)[2]
A_min_late <- exp(coef(fit_valleys_late)[1])
eta_min_late <- -coef(fit_valleys_late)[2]

# print results
cat("For time < 100:\n")
cat("A_max_early:", A_max_early, "\n")
cat("eta_max_early:", eta_max_early, "\n")
cat("A_min_early:", A_min_early, "\n")
cat("eta_min_early:", eta_min_early, "\n\n")

cat("For time > 100:\n")
cat("A_max_late:", A_max_late, "\n")
cat("eta_max_late:", eta_max_late, "\n")
cat("A_min_late:", A_min_late, "\n")
cat("eta_min_late:", eta_min_late, "\n")

```
#plot curve to test 
```{r}
# Generate exponential curves for early and late segments
max_exp_curve_early <- A_max_early * exp(-eta_max_early * time_early)
min_exp_curve_early <- A_min_early * exp(-eta_min_early * time_early)

max_exp_curve_late <- A_max_late * exp(-eta_max_late * time_late)
min_exp_curve_late <- A_min_late * exp(-eta_min_late * time_late)

# Convert data to data frames for ggplot
df_exp_curve_early <- data.frame(Time = time_early, MaxExpCurve = max_exp_curve_early, MinExpCurve = min_exp_curve_early)
df_exp_curve_late <- data.frame(Time = time_late, MaxExpCurve = max_exp_curve_late, MinExpCurve = min_exp_curve_late)

# Peaks and valleys of the original data
df_peaks <- data.frame(Time = t_peaks, Value = y_peaks)
df_valleys <- data.frame(Time = t_valleys, Value = y_valleys)

# Plot the original data and the fitted exponential curves
ggplot(df_MC, aes(x = Time, y = X1)) +
  geom_line(color = "blue", size = 1) +  # original data
  geom_line(data = df_exp_curve_early, aes(x = Time, y = MaxExpCurve), color = "purple", size = 1, linetype = "dashed") +  # Early segment max exponential curve
  geom_line(data = df_exp_curve_early, aes(x = Time, y = MinExpCurve), color = "yellow", size = 1, linetype = "dashed") +  # Early segment min exponential curve
  geom_line(data = df_exp_curve_late, aes(x = Time, y = MaxExpCurve), color = "purple", size = 1, linetype = "dotted") +  # late segment min exponential curve
  geom_line(data = df_exp_curve_late, aes(x = Time, y = MinExpCurve), color = "yellow", size = 1, linetype = "dotted") +  # late segment min exponential curve
  geom_point(data = df_peaks, aes(x = Time, y = Value), color = "red", size = 2, shape = 17) +  # Red points marking peaks
  geom_point(data = df_valleys, aes(x = Time, y = Value), color = "green", size = 2, shape = 18) +  # Green points marking valleys
  labs(title = "X1 Data with Early and Late Exponential Curve Fitting",
       x = "Time",
       y = "X1(t)") +
  theme_minimal() +
  scale_y_continuous(limits = c(0, max(X1) * 1.1))  # Adjust y-axis range to better display data

```

now calculate omega for time <100 and time>=100
```{r}
# For time < 100
# Calculate Delta and omega for early segment (time < 100)
Delta_early <- t_peaks_early[2] - t_peaks_early[1]  # Assuming the first two peaks are used
omega_early <- 2 * pi / Delta_early

# For time >= 100
# Calculate Delta and omega for late segment (time >= 100)
Delta_late <- t_peaks_late[2] - t_peaks_late[1]  # Assuming the first two peaks in the late segment are used
omega_late <- 2 * pi / Delta_late

# Print results
cat("For time < 100:\n")
cat("Delta_early:", Delta_early, "\n")
cat("omega_early:", omega_early, "\n\n")

cat("For time >= 100:\n")
cat("Delta_late:", Delta_late, "\n")
cat("omega_late:", omega_late, "\n")

```


then we should try to use initial parameters  fit  in nls and find mle
```{r}
#initial meeting model function
asym_model_function <- function(t, A_max, eta_max, A_min, eta_min, omega, phi_j) {
  
  A_max_t <- A_max * exp(-eta_max * t)
  A_min_t <- A_min * exp(-eta_min * t)
  
  result <- 0.5 * (A_max_t + A_min_t) + 0.5 * (A_max_t - A_min_t) * cos(omega * t + phi_j)
  return(result)
}

```




Detrending the Data (Exponential Method)
```{r}
# # Define detrending function using exponential method
# detrend_exponential <- function(time, data_column) {
#   # Fit an exponential model
#   trend_model_exp <- lm(log(data_column) ~ time)
#   
#   # Detrend the data by subtracting the fitted exponential trend
#   detrended_data <- data_column - exp(fitted(trend_model_exp))
#   
#   return(detrended_data)
# }
# 
# # Apply detrending to each column
# detrended_X1 <- detrend_exponential(df_MC$Time, df_MC$X1)
# detrended_X2 <- detrend_exponential(df_MC$Time, df_MC$X2)
# detrended_X3 <- detrend_exponential(df_MC$Time, df_MC$X3)
# detrended_X4 <- detrend_exponential(df_MC$Time, df_MC$X4)
# 
# # Plot detrended data
# plot(df_MC$Time, detrended_X1, type = "l", col = "blue", xlab = "Time", ylab = "Detrended Values", 
#      main = "Detrended X1, X2, X3, X4 (Exponential)")
# lines(df_MC$Time, detrended_X2, type = "l", col = "green")
# lines(df_MC$Time, detrended_X3, type = "l", col = "red")
# lines(df_MC$Time, detrended_X4, type = "l", col = "purple")
# 
# legend("topright", legend = c("X1", "X2", "X3", "X4"), col = c("blue", "green", "red", "purple"), lty = 1, cex = 0.8)

```
Spline Smooth
```{r}
# # Spline smoothing for detrended data
# spline_detrend_X1 <- predict(smooth.spline(df_MC$Time, detrended_X1))
# spline_detrend_X2 <- predict(smooth.spline(df_MC$Time, detrended_X2))
# spline_detrend_X3 <- predict(smooth.spline(df_MC$Time, detrended_X3))
# spline_detrend_X4 <- predict(smooth.spline(df_MC$Time, detrended_X4))
# 
# # Plot the spline-smoothed data
# plot(df_MC$Time, spline_detrend_X1$y, type = "l", col = "red", lwd = 2, 
#      xlab = "Time", ylab = "Values", main = "Detrended Splines", 
#      ylim = range(c(spline_detrend_X1$y, spline_detrend_X2$y, spline_detrend_X3$y, spline_detrend_X4)))
# lines(df_MC$Time, spline_detrend_X2$y, col = "blue", lwd = 2)
# lines(df_MC$Time, spline_detrend_X3$y, col = "green", lwd = 2)
# lines(df_MC$Time, spline_detrend_X4$y, col = "purple", lwd = 2)
# 
# legend("topright", legend = c("spline_detrend_X1", "spline_detrend_X2", "spline_detrend_X3", "spline_detrend_X4"), 
#        col = c("red", "blue", "green", "purple"), lwd = 2)

```
Nonlinear Least Squares Model Fitting with Weights
##time specific function

```{r}
# Define the model function
model_function <- function(t, A, gamma, omega, phi, y_shift) {
  A * exp(-gamma * t ) * cos(omega * t + phi) + y_shift
}
# Function to calculate variance at each time point
calculate_variance <- function(replicates) {
  n <- nrow(replicates)
  mean_vals <- rowMeans(replicates)  # Mean for each time point
  variance <- rowSums((replicates - mean_vals)^2) / (n - 1)  # Variance formula
  return(variance)
}
# Weighted nonlinear least squares estimation
weighted_nls <- function(t, replicates, initial_params) {
  variances <- calculate_variance(replicates)  # Variance at each time point

  # Fit the model using nlsLM
  fit <- nlsLM(y ~ model_function(t, A, gamma, omega, phi, y_shift),
               start = initial_params,
               data = list(t = t, y = rowMeans(replicates)),
               weights = 1 / variances)  # Weight by the inverse of variances
  return(fit)
}
# Use data from replicates for fitting
replicates <- df_MC[, c("X1", "X2", "X3", "X4")]
# Calculate variance
variances <- calculate_variance(replicates)
hist(variances, main = "Histogram of Variances",
     xlab = "Variance",
     ylab = "Frequency",
     col = "lightblue",
     border = "black")
# Initial parameter estimates

```







